Supermartingales and Termination Analysis of Statistical PCF

### 2020-05-28 probabilisticProgrammingMartingales.tex
Weaken the statement and correct the proof for antitone strict supermartingales.

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ adapted to filtration $(\calF_n)_{n \in \omega}$ is \emph{antitone-convex strict} if $Y_n \geq 0$, and there exists an antitone and convex measurable function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - (\epsilon \circ Y_n) \cdot 1_{\set{Y_n > 0}}
\]
%We say that the supermartingale is \emph{antitone-convex strict} if $\epsilon$ is further assumed to be convex.
\end{definition}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \in \omega}$ be a antitone-convex strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and define stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$. 
Then 
\begin{enumerate}
\item $T < \infty$ almost surely, and 
\item $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$.
\end{enumerate}
\end{theorem}

\begin{proof}
\emph{Claim}: for all $n \in \omega$,
\(
\expect{Y_n} \leq \expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big).
\)
We prove the claim by induction on $n$.
The base case is trivial.
For the inductive case, we have
\iffalse
\everymath{\displaystyle}
\begin{calculation}
\int_A\nu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)  
  \step{decomposition of $\nu = \kact{\mu }{ f}$} 
\int_A\mu(x,\dif y)\;f(x,y)\cdot \frac{\dif \nu'}{\dif\mu}(x,y)
  \step{$f=1$ outside $\infty[\nu]$} 
\int_A\mu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)
  \step{RN Theorem for $\sigma$-finite measures}
\int_A\nu'(x,\dif y)
  \step{definition of integral}
\nu'(x,A).
\end{calculation}
\fi
{%\everymath{displaystyle}
\begin{calculation}
\expect{Y_{n+1}}
  \step[=]{definition of conditional expectation}
\expect{\expect{Y_{n+1} \mid \calF_n}}
  \step[\leq]{linearity of expectation}
\expect{Y_n} - \int \epsilon \circ Y_n \cdot 1_{\set{Y_n > 0}} \, \dif \mu
\; = \;
\expect{Y_n} - \int_{\set{Y_n > 0}} \epsilon \circ Y_n \, \dif \mu
  \step[\leq]{convexity of $\epsilon$, and Jensen's inequality}
\expect{Y_n} - \epsilon \big(\int_{\set{Y_n > 0}} Y_n \, \dif \mu\big)
  \step[=]{by assumption $Y_n \geq 0$}
\expect{Y_n} - \epsilon (\expect{Y_n})
  \step[\leq]{$\mathbb{P}[Y_n > 0] \leq 1$}
\expect{Y_n} - \epsilon (\expect{Y_n}) \cdot \mathbb{P}[Y_n > 0]
  \step[\leq]{$\epsilon$ is antitone, and $\epsilon_0 = \expect{Y_0} \geq \expect{Y_n}$}
\expect{Y_n} - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
  \step[\leq]{induction hypothesis}
%\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big) - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
%\step[\leq]{}
\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
\end{calculation}
}

\iffalse
\begin{align*}
\expect{Y_{n+1}}
&=
\expect{\expect{Y_{n+1} \mid \calF_n}}\\
&\leq
\expect{Y_n} - \expect{\epsilon \circ Y_n} \cdot \mathbb{P}[Y_n > 0]\\
&\leq
\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big)
- \expect{\epsilon \circ Y_n} \cdot \mathbb{P}[Y_n > 0]\\
&\leq
\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
\end{align*}
\fi

It follows from the claim that $\sum_{i=0}^{\infty} \mathbb{P}[Y_i > 0]$ converges; hence $\lim_{n \to \infty} \mathbb{P}[Y_n > 0] = 0$ and so $\mathbb{P}[T < \infty] = 1$.

For 2, it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=1} \mathbb{P}[T \geq i]$.
\end{proof}

>>>>

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
Then $T < \infty$ almost surely.
\akr{It's not actually true that $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$. This is fortunate because it means that a rankability theorem based on antitone strict supermartingales actually has a chance of being stronger than the existing one (which was pretty much complete for $\tY$-PAST terms already). In fact, for any almost surely finite stopping time, there is a corresponding antitone strict supermartingale. This seems promising for the method being conditionally complete (for terms where all reachable terms are AST).}
\end{theorem}

\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$, therefore
\begin{calculation}
\expect{Y_n \mid T > n}
  \step[=]{}
\frac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[=]{}
\frac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{$Y_n \geq 0$ always}
\frac{\expect{Y_n}}{\mathbb P[T > n]}
  \step[\leq]{$(Y_n)_n$ is a supermartingale}
\frac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \frac{\expect{Y_0}+1}{x \epsilon(\expect{Y_0} x^{-1})} \right \rceil$, then
\(
\mathbb P[T > B_x] \leq x.
\)
As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
Suppose for a contradiction that $\mathbb P[T > B_x] > x$, then for every $n \leq B_x$
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{definition and linearity of conditional expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
  \step[=]{definition of conditional expectation for a discrete variable}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon(\frac{\expect{Y_0}}{\mathbb P[T > n]})
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x\ \epsilon(\expect{Y_0} x^{-1})
\end{calculation}
therefore, by a telescoping sum, $\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0 - B_x x \epsilon(\expect{Y_0} x^{-1})} \leq -1 < 0$, which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.

\end{proof}
